A thermodynamic view of dusty protoplanetary disks: axisymmetric stability

### 2016-08-18 model.tex
minor changes

\begin{align}
  &\frac{D\rho}{Dt} = -\rho\nabla\cdot\bm{v}, \label{masseq}\\ 
  &\frac{D\bm{v}}{Dt} = - \nabla\Phi - \frac{1}{\rho}\nabla  P, \label{momeq}\\ 
  &\frac{D\tepsilon}{Dt} = -\frac{1}{\rho} \nabla \cdot \left(\tepsilon 
  \tstop \nabla P \right),\label{dusteq}  
\end{align}
where $D/Dt \equiv \p_t + \bm{v}\cdot\nabla$ is the Lagragian
derivative; 
\begin{align}
  \rho \equiv \rhog + \rhod
\end{align}
is the total density with $\rhog$ and $\rhod$ being the gas and dust
density, respectively; 
\begin{align}
  \bm{v} \equiv \frac{\rhog\bm{v}_\mathrm{g} + 
    \rhod\bm{v}_\mathrm{d}}{\rho}
\end{align}
is the center-of-mass velocity with $\bm{v}_\mathrm{g}$ and
$\bm{v}_\mathrm{d}$ being the gas and dust velocities, respectively; 
and 
\begin{align}
  \tepsilon \equiv \frac{\rhod}{\rho}  = \frac{\epsilon}{1+\epsilon} 
\end{align}
is the dust-fraction and $\epsilon=\rhod/\rhog$ is the usual
dust-to-gas ratio. The gas pressure $P$ is given by the chosen 
equation of state described below. The external potential $\Phi$ is
that for a star of mass $M_*$ at the origin 
\begin{align}\label{thin_disk_potential}
  \Phi(r,z) =-\frac{GM_*}{\sqrt{r^2 + z^2}}\simeq
  -\frac{GM_*}{r}\left(1 - \frac{z^2}{2r^2}\right), 
\end{align}
where $G$ is the gravitational constant, and $(r,\phi, z)$ are
cylindrical co-ordinates centered on the star. 
The second equality is the 
thin-disk approximation for the disk potential, appropriate for
$|z|\ll r$. We shall adopt this approximate potential in order to
obtain analytic expressions for the disk equilibria.

@@

\subsection{Effective energy equation}
We show that for a prescribed temperature distribution, the mixture
obeys an evolutionary energy equation, due to the advection of the
dust-fraction. Eliminate $\tepsilon$ in favor of $P$, 
\begin{align*}
  \tepsilon = 1 - \frac{P}{c_s^2(r,z)\rho}, 
\end{align*}
then Eq. \ref{dusteq} becomes
\begin{align}
\frac{DP}{Dt} &= - P \nabla\cdot\bm{v} + P\bm{v}\cdot\nabla\ln{c_s^2}
                + \mathcal{C},  \label{eff_energy} \\
\mathcal{C}&\equiv c_s^2 \nabla\cdot\left[\tstop\left(1 -
  \frac{P}{c_s^2\rho}\right)\nabla 
  P\right].
\end{align}
Eq. \ref{eff_energy} is the energy equation for an ideal gas of adiabatic index
$\gamma=1$, i.e. isothermal evolution, but now with additional energy source 
terms due to the imposed temperature profile via $\nabla c_s^2$, as
well as dust-gas drag via $\mathcal{C}$. 
%In this form, the function $\mathcal{C}$ can be interpreted
%as a cooling term.    
If we denote 
\begin{align}
  \bm{F} \equiv  - \frac{\nabla P}{\rho}
\end{align}
for the pressure forces, then
\begin{align*}
  \mathcal{C} = - c_s^2\nabla \left( \tepsilon \tstop \rho \bm{F}
  \right), 
\end{align*}
which is in the same form as radiative diffusion. In protoplanetary
disks the heat flux, proportional to $\bm{F}$, is directly radially
outwards and vertically upwards. This is simply a reflection of
particle drift (inwards and downwards). Particle flux into a region contributes
to `cooling' of that region because the effective temperature is
lowered.

@@

\subsubsection{Possible instability at dust edges}
Notice the left-hand-side of Eq. \ref{stability_est2} is a quadratic in
$\p_r\epsilon$. It is possible to
violate this inequality for sufficiently large (in magnitude) radial
gradients  in the dust-to-gas ratio, 
\begin{align}
  \frac{\p\ln{\epsilon}}{\p\ln{r}} > S_+ \quad \mathrm{or} \quad 
  \frac{\p\ln{\epsilon}}{\p\ln{r}} < S_-,
\end{align}
\emph{for instability}, where
\begin{align}\label{spm}
S_\pm = \frac{1}{2}\frac{H_\mathrm{g}^2}{\Htilde^2} 
  \left[
  \frac{\p\ln{\rhog}}{\p\ln{r}} \pm 
  \sqrt{
  \left(\frac{\p\ln{\rhog}}{\p\ln{r}}\right)^2 + 
  4 \frac{\Htilde^2}{H_\mathrm{g}^2}
  \frac{\left(1+\epsilon\right)^2}{\epsilon h_\mathrm{g}^2}
  }
  \,\right]. 
\end{align} 
In typical accretion disks where $\p_r\rhog<0$, it is easier to
achieve instability at a given radius for increasing dust-to-gas
ratios in outwards ($\p_r\epsilon > 0$), and vice versa.

@@

The full energy equation is 
\begin{align*}
  \frac{DP}{Dt} = \frac{P}{\rho}\frac{D\rho}{D t} + \mathcal{C}. 
\end{align*}
Dust-drag causes a phase difference between the 
pressure and density evolution of a fluid element. Now, consider a 
fluid element at maximum density during an oscillation cycle, 
so that $D\rho/ D t = 0$. If at this point dust-gas drag (or particle flux) provides an
effective heating to the fluid element ($\mathcal{C}>0$), then it will 
experience increasing pressure, $DP/D t>0$. It attains 
pressure maximum \emph{after} reaching density maximum. This implies 
pressure maximizes while the parcel is expanding: it will do 
positive work  --- the area enclosed by its path in the `$P$-$V$' plane  is 
positive. This will increase the oscillation amplitude.

>>>>

\begin{align}
  &\frac{D\rho}{Dt} = -\rho\nabla\cdot\bm{v}, \label{masseq}\\ 
  &\frac{D\bm{v}}{Dt} = - \nabla\Phi - \frac{1}{\rho}\nabla  P, \label{momeq}\\ 
  &\frac{D\tepsilon}{Dt} = -\frac{1}{\rho} \nabla \cdot \left(\tepsilon 
  \tstop \nabla P \right),\label{dusteq}  
\end{align}
where $D/Dt \equiv \p_t + \bm{v}\cdot\nabla$ is the Lagragian
derivative; 
\begin{align}
  \rho \equiv \rhog + \rhod
\end{align}
is the total density with $\rhog$ and $\rhod$ being the gas and dust
density, respectively; 
\begin{align}
  \bm{v} \equiv \frac{\rhog\bm{v}_\mathrm{g} + 
    \rhod\bm{v}_\mathrm{d}}{\rho}
\end{align}
is the center-of-mass velocity with $\bm{v}_\mathrm{g}$ and
$\bm{v}_\mathrm{d}$ being the gas and dust velocities, respectively; 
and 
\begin{align}
  \tepsilon \equiv \frac{\rhod}{\rho}  = \frac{\epsilon}{1+\epsilon} 
\end{align}
is the dust-fraction and $\epsilon=\rhod/\rhog$ is the usual
dust-to-gas ratio. The gas pressure $P$ is given by the chosen 
equation of state described below. Note that Eq. \ref{dusteq} is
equivalent to Eq. 46 of \cite{jacquet11}.

The external potential $\Phi$ is 
that for a star of mass $M_*$, 
\begin{align}\label{thin_disk_potential}
  \Phi(r,z) =-\frac{GM_*}{\sqrt{r^2 + z^2}}\simeq
  -\frac{GM_*}{r}\left(1 - \frac{z^2}{2r^2}\right), 
\end{align}
where $G$ is the gravitational constant, and $(r,\phi, z)$ are
cylindrical co-ordinates centered on the star. 
The second equality is the 
thin-disk approximation for the disk potential, appropriate for
$|z|\ll r$. We shall adopt this approximate potential in order to
obtain analytic expressions for the disk equilibria.

@@

\subsection{Effective energy equation}
We show that for a prescribed temperature distribution, the mixture
obeys an evolutionary energy equation, due to the advection of the
dust-fraction. Eliminate $\tepsilon$ in favor of $P$, 
\begin{align*}
  \tepsilon = 1 - \frac{P}{c_s^2(r,z)\rho}, 
\end{align*}
then Eq. \ref{dusteq} becomes
\begin{align}
\frac{DP}{Dt} &= - P \nabla\cdot\bm{v} + P\bm{v}\cdot\nabla\ln{c_s^2}
                + \mathcal{C},  \label{eff_energy} \\
\mathcal{C}&\equiv c_s^2 \nabla\cdot\left[\tstop\left(1 -
  \frac{P}{c_s^2\rho}\right)\nabla 
  P\right].
\end{align}
Eq. \ref{eff_energy} is the energy equation for an ideal gas of adiabatic index
$\gamma=1$, i.e. isothermal evolution, but now with additional energy source 
terms due to the imposed temperature profile via $\nabla c_s^2$, as
well as dust-gas drag via $\mathcal{C}$. 
%In this form, the function $\mathcal{C}$ can be interpreted
%as a cooling term.    
If we denote 
\begin{align}
  \bm{F} \equiv  - \frac{\nabla P}{\rho}
\end{align}
for the pressure forces, then
\begin{align*}
  \mathcal{C} = - c_s^2\nabla \left( \tepsilon \tstop \rho \bm{F}
  \right), 
\end{align*}
which is in the same form as cooling by radiative diffusion. In protoplanetary
disks the heat flux, proportional to $\bm{F}$, is directly radially
outwards and vertically upwards. This is simply a reflection of
particle drift (inwards and downwards). Particle flux into a region contributes
to `cooling' of that region because the effective temperature is
lowered.

@@

\subsubsection{Instability at dust edges}
Notice the left-hand-side of Eq. \ref{stability_est2} is a quadratic in
$\p_r\epsilon$. It is possible to
violate this inequality for sufficiently large (in magnitude) radial
gradients  in the dust-to-gas ratio, 
\begin{align}
  \frac{\p\ln{\epsilon}}{\p\ln{r}} > S_+ \quad \mathrm{or} \quad 
  \frac{\p\ln{\epsilon}}{\p\ln{r}} < S_-,
\end{align}
\emph{for instability}, where
\begin{align}\label{spm}
S_\pm = \frac{1}{2}\frac{H_\mathrm{g}^2}{\Htilde^2} 
  \left[
  \frac{\p\ln{\rhog}}{\p\ln{r}} \pm 
  \sqrt{
  \left(\frac{\p\ln{\rhog}}{\p\ln{r}}\right)^2 + 
  4 \frac{\Htilde^2}{H_\mathrm{g}^2}
  \frac{\left(1+\epsilon\right)^2}{\epsilon h_\mathrm{g}^2}
  }
  \,\right]. 
\end{align} 
In typical accretion disks where $\p_r\rhog<0$, it is easier to
achieve instability at a given radius for increasing dust-to-gas
ratios in outwards ($\p_r\epsilon > 0$), and vice versa.

@@

The full energy equation is 
\begin{align*}
  \frac{DP}{Dt} = \frac{P}{\rho}\frac{D\rho}{D t} + \mathcal{C}. 
\end{align*}
Dust-drag causes a phase difference between the 
pressure and density evolution of a fluid element. Now, consider a 
fluid element at maximum density during an oscillation cycle, 
so that $D\rho/ D t = 0$. If at this point dust-gas drag provides an
effective heating to the fluid element ($\mathcal{C}>0$, i.e. a particle
flux \emph{away} from it), then it will 
experience increasing pressure, $DP/D t>0$. It will attain  
pressure maximum \emph{after} reaching density maximum. This implies 
pressure maximizes while the parcel is expanding: it will do 
positive work  --- the area enclosed by the parcel's path in the `$P$-$V$' plane  is 
positive. This will increase the oscillation amplitude.

### 2016-08-18 paper.tex
minor changes

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  This is analogous to the 
  conservation of entropy 
  following an adiabatic fluid.  We show that in the strong drag
  limit that an isothermal dusty gas indeed behaves like an  
  adiabatic pure gas, and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this equivalence 
  to 
%\begin{inparaenum}[1)]
%\item
%  identify the entropy of isothermal dusty gas; and 
%\item 
  generalize previous, well-known gaseous instabilities
  to dusty protoplanetary disks.
%\end{inparaenum}
%This includes the axisymmetric stability of strictly isothermal dusty
%gas, allowing us to constrain the sharpness of 
%dust edges that can persist in protoplanetary disks. 
We show that strictly isothermal, non-self-gravitating, axisymmetric
disks perfectly coupled to dust are generally stable, however thin the
dust layer is.  %except possibily in the presence of sharp 
%edges in radial dependence of the dust-to-gas ratio. 
We also develop 
the dusty vertical shear instability (VSI) in locally isothermal
disks. We find that  dust-loading reduces VSI growth rates, and
preferentially stabilizes VSI modes with small radial 
wavelengths, as well as `surface modes' concentrated near the
disk boundaries. These results can be understood in terms of 
the buoyancy frequency of a isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We briefly discuss potential thermodynamical interpretations of 
dust-gas drag dynamics, such as the streaming instability in
protoplanetary disks as a manifestation of pulsational instabilities
in stars.  
\end{abstract}

>>>>

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  This is analogous to the 
  conservation of entropy 
  following an adiabatic fluid.  We show that in the strong drag
  limit that an isothermal dusty gas indeed behaves like an  
  adiabatic pure gas; and that finite dust-gas coupling may be
  considered as an effective 
  heating/cooling. We exploit this equivalence 
  to 
%\begin{inparaenum}[1)]
%\item
%  identify the entropy of isothermal dusty gas; and 
%\item 
  generalize previous, well-known gaseous instabilities
  to dusty protoplanetary disks.
%\end{inparaenum}
%This includes the axisymmetric stability of strictly isothermal dusty
%gas, allowing us to constrain the sharpness of 
%dust edges that can persist in protoplanetary disks. 
We show that strictly isothermal, non-self-gravitating, axisymmetric
disks perfectly coupled to dust are generally stable, however thin the
dust layer is.  %except possibily in the presence of sharp 
%edges in radial dependence of the dust-to-gas ratio. 
We also develop 
the dusty vertical shear instability (VSI) in locally isothermal
disks. We find that  dust-loading reduces VSI growth rates, and
preferentially stabilizes VSI modes with small radial 
wavelengths, as well as `surface modes' concentrated near the
disk boundaries. These results can be understood in terms of 
the buoyancy frequency of a isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We briefly discuss thermodynamical interpretations of 
dust-gas drag dynamics, such as the streaming instability in
protoplanetary disks as a manifestation of pulsational instabilities
in stars.  
\end{abstract}
